---
title: "Data Sources"
page-layout: article
execute:
  warning: false
  message: false
---

This page describes the French corpora used in the package vignettes and available in the `data-raw/corpus-data/` directory as both RDA and Parquet files.

## Overview

The package uses two main corpus collections:

1. **Chambers–Le Baron Corpus** (academic French, 159 documents)
2. **French Register Corpus** (multi-register, 9,310 documents across 6 registers)

## Chambers–Le Baron Corpus

**Source:** [Oxford Text Archive](https://ota.bodleian.ox.ac.uk/repository/xmlui/handle/20.500.12024/2527)

**Description:** A specialized corpus of French academic writing spanning 10 disciplines. This corpus is used in the [Chambers–Le Baron Walkthrough](vignettes/chambers-corpus.qmd) vignette to demonstrate disciplinary variation in French academic discourse.

**Composition:**

- **Documents:** 159 research articles
- **Total tokens:** 125,445
- **Disciplines:** Economics, History, Information and Communication, Law, Linguistics/Language Learning, Literature, Media/Culture, Philosophy, Social Anthropology, Social Sciences
- **Text type:** Published academic research articles

```{r}
#| echo: false
#| eval: true

library(dplyr)
library(gt)

# Discipline breakdown with token counts
tibble::tribble(
  ~Discipline, ~Documents, ~Tokens, ~`Mean tokens/doc`,
  "Economics", "16", "12,185", "762",
  "History", "13", "10,411", "801",
  "Information and Communication", "15", "11,639", "776",
  "Law", "19", "15,147", "797",
  "Linguistics/Language Learning", "15", "11,588", "773",
  "Literature", "20", "16,191", "810",
  "Media/Culture", "18", "14,594", "811",
  "Philosophy", "13", "10,268", "790",
  "Social Anthropology", "13", "10,279", "791",
  "Social Sciences", "17", "13,143", "773"
) |>
  gt() |>
  tab_header(
    title = "Chambers–Le Baron Corpus by Discipline"
  ) |>
  cols_align(
    align = "left",
    columns = Discipline
  ) |>
  cols_align(
    align = "right",
    columns = c(Documents, Tokens, `Mean tokens/doc`)
  )
```

**Files:**

- `french_corpus.rda` / `french_corpus.parquet`
- `fc_udpipe_parsed.rda` / `fc_spacy_parsed.rda` (pre-parsed versions in `_quarto/data/`)

## French Register Corpus

**Description:** A multi-register corpus compiled to validate feature extraction across diverse text types and demonstrate register variation in French. This corpus is used in the [Cross-Parser Validation](vignettes/parser-validation.qmd) vignette.

**Composition:**

- **Total documents:** 9,310
- **Total words:** ~1.4 million
- **Registers:** 6 (academic, fiction, news, spoken, twitter, wikipedia)

```{r}
#| echo: false
#| eval: true

library(dplyr)
library(gt)

# Register breakdown with token counts
tibble::tribble(
  ~Register, ~Documents, ~Tokens, ~`Mean tokens/doc`, ~Description,
  "Academic", "159", "125,445", "789", "Research articles (Chambers–Le Baron corpus)",
  "Fiction", "46", "397,777", "8,647", "Literary texts sampled by line chunks (150 lines per file)",
  "News", "503", "362,136", "720", "Journalistic articles (Chambers-Rostand corpus)",
  "Spoken", "87", "160,026", "1,839", "Oral narratives (French Oral Narrative Corpus)",
  "Twitter", "7,197", "199,263", "28", "French tweets (excluded from MDA due to extreme brevity)",
  "Wikipedia", "1,318", "370,282", "281", "Wikipedia articles"
) |>
  gt() |>
  tab_header(
    title = "French Register Corpus Composition"
  ) |>
  cols_align(
    align = "left",
    columns = c(Register, Description)
  ) |>
  cols_align(
    align = "right",
    columns = c(Documents, Tokens, `Mean tokens/doc`)
  )
```

### Register Details

#### Academic (159 documents)

Subset of the Chambers–Le Baron corpus representing formal academic prose across multiple disciplines.

#### Fiction (46 documents)

**Sampling method:** Line-based sampling from 100 French literary texts

- Extracted 150-line chunks from diverse fiction sources
- Ensures variety across authors and periods
- Avoids over-representation from single long novels

#### News (503 documents)

**Source:** [Chambers-Rostand Corpus of Journalistic French](https://ota.bodleian.ox.ac.uk/repository/xmlui/handle/20.500.12024/2491)

Contemporary French journalism covering various topics and publication outlets.

#### Spoken (87 documents)

**Source:** [French Oral Narrative Corpus](https://ota.bodleian.ox.ac.uk/repository/xmlui/handle/20.500.12024/2555)

Transcribed oral narratives representing conversational French. Includes features typical of spoken interaction: filled pauses, discourse particles, informal contractions.

#### Twitter (7,197 documents)

**Note:** Excluded from multidimensional analyses due to extreme brevity (mean 22 words/document), which creates normalization artifacts when computing features per 1,000 words.

French tweets collected to represent informal digital communication.

#### Wikipedia (1,318 documents)

French Wikipedia articles representing informational but accessible encyclopedic prose. More formal than conversational registers but less technical than academic writing.

## Data Files

All corpora are stored in `data-raw/corpus-data/` in two formats:

**Parquet format** (recommended for large datasets):
```
french_corpus.parquet             # Chambers–Le Baron corpus
french_register_corpus.parquet    # Multi-register corpus
```

**RDA format** (R-native):
```
french_corpus.rda
french_register_corpus.rda
```

**Pre-parsed versions** for documentation rendering (in `_quarto/data/`):
```
fc_udpipe_parsed.rda    # Chambers–Le Baron parsed with UDPipe
fc_spacy_parsed.rda     # Chambers–Le Baron parsed with spaCy
fr_features_ud.rda      # Extracted features (UDPipe)
fr_features_spacy.rda   # Extracted features (spaCy)
```

## Usage in Vignettes

### Getting Started
Uses synthetic examples to demonstrate basic workflow without requiring corpus data.

### Cross-Parser Validation
Uses the **French Register Corpus** (excluding Twitter) to compare UDPipe and spaCy parsers across 5 registers: academic, fiction, news, spoken, wikipedia.

### Chambers–Le Baron Walkthrough
Uses the **Chambers–Le Baron Corpus** to analyze disciplinary variation in French academic writing.

## Citations

When using these corpora in research, please cite:

**Chambers–Le Baron Corpus:**
> Chambers, A., & Le Baron, M. (2007). *A Corpus of Academic French*. Oxford Text Archive. [https://ota.bodleian.ox.ac.uk/repository/xmlui/handle/20.500.12024/2527](https://ota.bodleian.ox.ac.uk/repository/xmlui/handle/20.500.12024/2527)

**French Oral Narrative Corpus:**
> Université catholique de Louvain. *French Oral Narrative Corpus*. Oxford Text Archive. [https://ota.bodleian.ox.ac.uk/repository/xmlui/handle/20.500.12024/2555](https://ota.bodleian.ox.ac.uk/repository/xmlui/handle/20.500.12024/2555)

**Chambers-Rostand Corpus:**
> Chambers, A., & Rostand, C. *Chambers-Rostand Corpus of Journalistic French*. Oxford Text Archive. [https://ota.bodleian.ox.ac.uk/repository/xmlui/handle/20.500.12024/2491](https://ota.bodleian.ox.ac.uk/repository/xmlui/handle/20.500.12024/2491)

## Accessing the Data

The corpora are **not distributed with the package** due to size and licensing considerations. They are used for documentation and validation purposes, with pre-computed feature extractions stored in `_quarto/data/` for vignette rendering.

For researchers interested in replicating analyses:

1. Download corpora from Oxford Text Archive (links above)
2. Place raw text files in `data-raw/corpus-data/`
3. Use parsing scripts in `data-raw/` to regenerate parsed versions
4. Extract features using `biber()` function
