---
title: "Cross-Parser Validation of French Register Dimensions"
format: html
execute:
  warning: false
  message: false
---

## Introduction

This vignette demonstrates the robustness of multidimensional analysis (MDA) for French register variation by comparing results across two different NLP parsers: **spaCy** and **UDPipe**. Despite different parsing algorithms and training data, both models produce remarkably consistent dimensional structures.

We also compare the French dimensional structure to Biber's (1988) original English dimensions, showing cross-linguistic alignment in the fundamental distinction between **Interactional** and **Information Production** registers.

::: {.callout-note}
## Data
This vignette uses the **French Register Corpus** (1.4M tokens, 6 registers). For detailed corpus composition and token counts by register, see [Data Sources](../data-sources.qmd).
:::

## Cross-Parser Methodology

We extracted 66 linguistic features from the same French register corpus using both parsers, then performed parallel MDAs to test whether the dimensional structure is stable across NLP pipelines.

### Loading the Data

Both feature sets were pre-computed and normalized to counts per 1,000 words. The feature extraction workflow is shown below (not executed during rendering):

```{r}
#| eval: false

# Feature extraction workflow (not executed - data loaded from saved files)
fr_features_spacy <- biber(fr_spacy_parsed, measure = "none", normalize = TRUE) |>
  left_join(french_register_corpus |> select(doc_id, register), by = "doc_id") |>
  select(doc_id, register, everything()) |>
  mutate(register = as.factor(register))

fr_features_ud <- biber(fr_udpipe_parsed, measure = "none", normalize = TRUE) |>
  left_join(french_register_corpus |> select(doc_id, register), by = "doc_id") |>
  select(doc_id, register, everything()) |>
  mutate(register = as.factor(register))
```

We load the pre-extracted features from both parsers:

```{r}
#| label: check-data
#| include: false

# Check if data files are available
data_available <- file.exists("../data/fr_features_spacy.rda") && 
                  file.exists("../data/fr_features_ud.rda")
```

```{r}
#| eval: !expr data_available

library(pseudobibeR.fr)
library(mda.biber)
library(ggplot2)
library(tidyr)
library(dplyr)
library(stringr)
library(gt)
library(broom)
library(gtsummary)
library(tibble)

# Load pre-extracted features from both parsers
load("../data/fr_features_spacy.rda")
load("../data/fr_features_ud.rda")

# Remove twitter due to extreme brevity (mean 22 words/doc)
fr_features_spacy <- fr_features_spacy %>% filter(register != "twitter")
fr_features_ud <- fr_features_ud %>% filter(register != "twitter")

cat("spaCy corpus:", nrow(fr_features_spacy), "documents\n")
cat("UDPipe corpus:", nrow(fr_features_ud), "documents\n")
```

::: {.callout-note}
## Data Availability
This vignette requires pre-computed feature data files. If you're rendering locally without these files, the code chunks will not execute but are shown for reference.
:::

## Scree Plots: Determining Dimensionality

The scree plots show the proportion of variance explained by each factor:

```{r}
#| eval: !expr data_available
#| layout-ncol: 2
#| fig-cap: 
#|   - "spaCy eigenvalues"
#|   - "UDPipe eigenvalues"

screeplot_mda(fr_features_spacy)
screeplot_mda(fr_features_ud)
```

Both parsers suggest a **3-factor solution** with a clear elbow at Factor 4. The eigenvalue profiles are nearly identical, indicating stable dimensionality.

## Factor Loadings Comparison

We extract 3 factors using a minimum correlation of 0.2 and loading threshold of 0.35:

```{r}
#| eval: !expr data_available

loadings_spacy <- mda_loadings(
  obs_by_group = fr_features_spacy,
  n_factors = 3,
  cor_min = 0.2,
  threshold = 0.35
)

loadings_ud <- mda_loadings(
  obs_by_group = fr_features_ud,
  n_factors = 3,
  cor_min = 0.2,
  threshold = 0.35
)
```

### spaCy Loadings

```{r}
#| eval: !expr data_available

attr(loadings_spacy, 'loadings') |>
  tibble::rownames_to_column("Feature") |>
  gt() |>
  fmt_number(
    columns = where(is.numeric),
    decimals = 2
  ) |>
  tab_header(
    title = "Factor Loadings: spaCy Parser"
  )
```

### UDPipe Loadings

```{r}
#| eval: !expr data_available

attr(loadings_ud, 'loadings') |>
  tibble::rownames_to_column("Feature") |>
  gt() |>
  fmt_number(
    columns = where(is.numeric),
    decimals = 2
  ) |>
  tab_header(
    title = "Factor Loadings: UDPipe Parser"
  )
```

### Interpreting the Loadings

Both parsers reveal remarkably consistent factor structures:

**Factor 1** represents an **Interactional vs. Information Production** dimension. The positive pole is characterized by:

- **Pronouns**: Second person (spaCy: 0.83, UDPipe: 0.73), first person (spaCy: 0.78, UDPipe: 0.59), third person (UDPipe: 0.69)
- **Interactive features**: WH-questions (0.56/0.54), discourse particles (0.44/0.51), clausal coordination (0.58/0.55)
- **Deixis**: Demonstrative pronouns (spaCy: 0.74), indefinite pronouns (0.41/0.47)

The negative pole features:

- **Nominal style**: Other nouns (−0.86/−0.63), nominalizations (−0.49/−0.50)
- **Structural complexity**: Prepositions (−0.51/−0.48), mean word length (−0.33/−0.33)
- **Non-finite forms**: Phrasal coordination (−0.19/−0.20), past participle whiz-deletions (−0.21/−0.19)

**Factor 2** shows **explicit stance and subordination** features with high loadings on adverbs (0.64/0.51), subordinators (0.50/0.69), infinitives (UDPipe: 0.66), and complement clauses.

**Factor 3** differs slightly between parsers but captures **perfect aspect** (UDPipe: 0.92) and **passive constructions** (UDPipe: agentless passives 0.71).

The cross-parser consistency is striking: key features maintain similar loading patterns, and the fundamental Interactional vs. Informational contrast emerges identically from both NLP pipelines.

## Factor 1: Interactional vs. Information Production

Factor 1 captures the primary dimension of register variation in both analyses. The stick plots show register positions along this dimension:

```{r}
#| eval: !expr data_available
#| layout-ncol: 2
#| fig-cap: 
#|   - "spaCy Factor 1"
#|   - "UDPipe Factor 1"

stickplot_mda(loadings_spacy, n_factor = 1)
stickplot_mda(loadings_ud, n_factor = 1)
```

### Heatmap Visualization

```{r}
#| eval: !expr data_available
#| layout-ncol: 2
#| fig-cap: 
#|   - "spaCy register clustering"
#|   - "UDPipe register clustering"

heatmap_mda(loadings_spacy, n_factor = 1)
heatmap_mda(loadings_ud, n_factor = 1)
```

## Statistical Validation

### ANOVA: Register Differences on Factor 1

Both parsers show highly significant register effects:

```{r}
#| eval: !expr data_available

f_aov_spacy <- aov(Factor1 ~ group, data = loadings_spacy)
f_aov_ud <- aov(Factor1 ~ group, data = loadings_ud)

bind_rows(
  broom::tidy(f_aov_spacy) |> mutate(Parser = "spaCy"),
  broom::tidy(f_aov_ud) |> mutate(Parser = "UDPipe")
) |>
  select(Parser, everything()) |>
  gt() |>
  fmt_number(
    columns = c(sumsq, meansq, statistic),
    decimals = 2
  ) |>
  fmt(
    columns = p.value,
    fns = function(x) {
      ifelse(x < 0.001, "< .001", sprintf("%.3f", x))
    }
  ) |>
  tab_header(
    title = "ANOVA: Factor 1 ~ Register"
  )
```

### Regression Models

Linear regression confirms strong register effects with both parsers. The UDPipe model achieves R² = 0.715 (Adjusted R² = 0.714), slightly higher than the spaCy model's R² = 0.703 (Adjusted R² = 0.702). Both models show highly significant F-statistics (p < 0.001), indicating that register explains approximately 70% of the variance in Factor 1 scores.

The coefficient patterns are nearly identical across parsers, with spoken and fiction registers at the positive (interactional) pole and academic, news, and Wikipedia at the negative (informational) pole. This cross-parser consistency validates the robustness of the dimensional structure.

```{r}
#| eval: !expr data_available

gtsummary::tbl_merge(
  list(
    gtsummary::tbl_regression(lm(Factor1 ~ group, data = loadings_ud)) |>
      gtsummary::add_glance_source_note(),
    gtsummary::tbl_regression(lm(Factor1 ~ group, data = loadings_spacy)) |>
      gtsummary::add_glance_source_note()
  ),
  tab_spanner = c("**UDPipe (R² = 0.715)**", "**spaCy (R² = 0.703)**")
) |>
  gtsummary::as_gt() |>
  tab_header(
    title = "Linear Models: Factor 1 ~ Register"
  )
```

## Cross-Linguistic Alignment

### Biber's Dimension 1 (English)

Biber's (1988) Dimension 1 contrasts:

**Positive pole (Involved/Interactional):**
- Private verbs (think, feel)
- First/second person pronouns
- Contractions
- Present tense
- Discourse particles

**Negative pole (Informational):**
- Nouns
- Long words
- Prepositions
- Attributive adjectives
- Past participles

### French Factor 1

Our French analysis replicates this fundamental distinction:

**Positive pole (Interactional):**
- Spoken and fiction registers (dialogue)
- First/second person pronouns
- Contractions (p'tit, m'sieur)
- Present tense
- "Que" deletion (informal complementizer omission)

**Negative pole (Informational):**
- Academic, news, and Wikipedia registers
- Nominalizations
- Prepositions
- Attributive adjectives
- Longer words

## Key Implementation Detail: Contraction Detection

A critical validation was ensuring that the **contraction detection** (f_59_contractions) worked consistently across parsers. French distinguishes:

- **Grammatical elisions** (mandatory): l' (le), d' (de), qu' (que) - articles, prepositions
- **True contractions** (informal): p'tit (petit), m'sieur (monsieur) - lexical words

We implemented POS-based filtering:
```r
# Exclude tokens with grammatical POS tags
elision_pos_tags <- c("DET", "ADP", "PRON", "ADV", "SCONJ")
contractions <- apostrophe_tokens %>%
  filter(!pos %in% elision_pos_tags)
```

Both spaCy and UDPipe assign identical POS tags to elisions, ensuring consistent counts:

- **spaCy**: l' (DET), d' (ADP), s' (PRON), n' (ADV), qu' (SCONJ)
- **UDPipe**: l' (DET), d' (ADP), s' (PRON), n' (ADV), qu' (SCONJ)

Both parsers correctly tag true contractions with lexical POS (ADJ, NOUN, PROPN), allowing them to pass the filter.

## Conclusions

1. **Parser robustness**: spaCy and UDPipe produce nearly identical MDA results, validating the feature extraction implementation

2. **Cross-linguistic universality**: French Factor 1 aligns with Biber's English Dimension 1, supporting the theory that Interactional vs. Information Production is a fundamental axis of register variation

3. **Implementation reliability**: POS-based feature detection (especially for contractions) works consistently across NLP pipelines

4. **Methodological transparency**: The `pseudobibeR.fr` package enables reproducible register analysis with multiple parser backends

## References

Biber, D. (1988). *Variation across speech and writing*. Cambridge University Press.

## Session Info

```{r}
sessionInfo()
```
